Extracting sunbeams from cucumbers: Reproducibility as a necessity for the future of archaeological science

### 2025-01-31 paper.qmd
minor edits

At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by a generous measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.

@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through to the reproducibility review so far. The most commonly used software is R, followed by Python. Results generated with proprietary or closed-source software are out of scope for reproducibility reviews. Several distinct types of analyses are well-represented in this sample, especially geometric morphometry, network statistics, and analyses using artificial intelligence or machine learning algorithms (this includes deep learning and neural networks). Most authors are sharing their code and data files via Zenodo, a non-profit generic research data repository hosted by CERN that accepts any file format and freely assigns all publicly available uploads a DOI to make the files easily and uniquely citable [@petersZenodoSpotlightTraditional2017]. In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories. GitHub, a commercial service owned by Microsoft is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability of Zenodo. Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.

Panels D and E of @fig-aer-summary summarise the common issues that resulted in irreproducible results. The most common issue was an incomplete compendium. This ranges from missing data files down to missing lines of code, and in most cases seems to be due to carelessness. Unspecified or under-specified dependencies is another common issue that prevents code from running. This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations, (e.g. dplyr for R or numpy for Python). If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change they way they behave as the developers update their package. Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.

>>>>

At the time of writing (Feburary 2025) I have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values obtained from a variety of harder and soft sciences.

@fig-aer-summary shows a summary of basic characteristics of the 25 articles that have been through the reproducibility review process so far. The most commonly used software is R, followed by Python. Results generated with proprietary or closed-source software are out of scope for reproducibility reviews. Several distinct types of analyses are well-represented in this sample, especially geometric morphometry, network statistics, and analyses using artificial intelligence or machine learning algorithms (this includes deep learning and neural networks). Most authors are sharing their code and data files via Zenodo, a non-profit generic research data repository hosted by CERN that accepts any file format and freely assigns all publicly available uploads a DOI to make the files easily and uniquely citable [@petersZenodoSpotlightTraditional2017]. In this same category of DOI-issuing, research-grade repositories is OSF (the Open Science Foundation), Figshare, and university repositories. GitHub, a commercial service owned by Microsoft is a code hosting platform that is convenient for collaboration, is also popular among JAS authors, but is a problematic choice because does not offer DOIs or the same commitments to long-term availability as Zenodo. Some authors attached their code and data as journal article supplementary files, but this is a poor choice for long-term availability because these files are typically renamed and converted to different formats during the article production process, making it difficult or impossible for a reader to combine the code and data to reproduce the results.

Panels D and E of @fig-aer-summary summarise the common issues that resulted in irreproducible results. The most common issue was an incomplete compendium. This ranges from missing data files down to missing lines of code. In most cases this can attributed to accidental carelessness, with the exception of two cases where data was unavailable due to licensing restrictions. Unspecified or under-specified dependencies is another common issue that prevents code from running. This refers to the software packages in addition to R or Python that an author used to do specialized analyses and visualisations, (e.g. dplyr for R or numpy for Python). If an author does not clearly specify the name and version number of the packages that they used for their analysis, it can be very time-consuming or impossible to correctly identify these because many packages have functions with similar names, and functions in any one package can change they way they behave as the developers update their package. Other reasons why papers failed the reproducibility review is that the paths to data files were incorrectly specified (likely a result of the author reorganising their compendium after completing their analysis, or omitting data files from the materials submitted for review), and errors returned by functions, which have diverse causes.
